Paper 1: Triangulating trends in inequality around the world from secondary databases

### 2023-02-09 PhD_pages/data_appendices/1_F_within_country_trends.qmd
Paper 1: reorganise app and initial drafting of paper

---
title: "Appendix 1.F: Analysing global trends in within-country inequality"
format: html
warning: false
filters:
  - shinylive

---
```{python}

from plotnine import *
import pandas as pd
import numpy as np

@@

This first block of code.....

@@

This second block of code we read in the data (this needs to be done in a different way in the shiny app...)

#  ---------------------------------
#  ---------- SECTION 2: Read in data -------------
#  ----------------------------------

In a third block, we set specifications for the data matching. This is different in the app because some parameters are controlled by the app.

```{python}
#  ---------------------------------
#  ---------- SECTION 3: Specifications -------------
#  ----------------------------------

# These parameters are not controlled in the app.
reference_col = 'Year'
group_by_col = 'Entity'

tolerance = 1
outlier_cut_off_upper = None

source_count_requirement = "all"
# source_count_requirement = "any"

reference_vals = [1990,2018]
max_dist_from_refs = 4
min_dist_between = 5

data_to_match = {
  "sources": ["PIP", "WID", "LIS"],
  "var_names": ["Top_10_share", "Top_10_share_pretax", "Top_10_share_equiv_disposable"]
  }

In this code block we run the code and prepare final plots and tables.

#  ---------------------------------
#  ---------- SECTION 4: Run data functions and produce plot -------------
#  ----------------------------------

@@

def prep_and_merge_multiple_vars(data_to_match, reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col):

@@

    source = data_to_match["sources"][i]
    var = data_to_match["var_names"][i]
    print(i)
    print(source)
    print(var)
    df = data[source]

@@

  ref_matches = pd.concat(matches, keys=source_var_keys)

  # Tidy up indexes
  ref_matches = ref_matches.reset_index()

  ref_matches = ref_matches.drop('level_1', axis=1)

  ref_matches = ref_matches\
    .rename(columns={"level_0": "source_var"})

  return ref_matches

def prep_plot_and_tables(
  data_to_match,
  reference_vals, 
  max_dist_from_refs,
  min_dist_between,
  reference_col,
  group_by_col,
  tolerance,
  outlier_cut_off_upper,
  source_count_requirement):

  # Other spec:
  region_col = 'region_alt'

  # ---- Prep data ---------

  plot_data = prep_and_merge_multiple_vars(
    data_to_match = data_to_match,
    reference_vals = reference_vals, 
    max_dist_from_refs = max_dist_from_refs,
    min_dist_between = min_dist_between,
    reference_col = reference_col,
    group_by_col = group_by_col
  )

@@

  # Apply source requirement (whether to include only observations with data from all sources)
  if source_count_requirement == "all":
    plot_data = plot_data.loc[plot_data['source_count'] == plot_data['source_count'].max()]

@@

  # ----- Prep plot ------

  # Set the coordinates of the shaded ribbon
  shaded_coord = pd.DataFrame({'x': [x_min_floor, x_max_ceiling, x_max_ceiling, x_min_floor], 
    'y': [x_min_floor-tolerance, x_max_ceiling-tolerance, x_max_ceiling+tolerance , x_min_floor+tolerance]})

  plot = (ggplot(plot_data
  , aes(x_axis, y_axis, alpha=0.5))
  + geom_point(aes(color=region_col, size='avg_pop', alpha = 0.6))
  + facet_wrap('~ source_var')
  + geom_polygon(aes(x='x', y='y', alpha = 0.2), data=shaded_coord)
  + theme_light()
  + theme(legend_position='none')
  )

  # ---- Prep summary table ----

@@

  # ----- Return outputs ------
  plot_and_tables = {
    "table_data": plot_data,
    "plot": plot,
    "summary": df_summary
  }

  return plot_and_tables

@@

```{python}
plot_and_tables = prep_plot_and_tables(
    data_to_match = data_to_match,
    reference_vals = reference_vals, 
    max_dist_from_refs = max_dist_from_refs,
    min_dist_between = min_dist_between,
    reference_col = reference_col,
    group_by_col = group_by_col,
    tolerance = tolerance,
    outlier_cut_off_upper = outlier_cut_off_upper,
    source_count_requirement = source_count_requirement
    )

plot = plot_and_tables['plot']
summary = plot_and_tables['summary']

@@

```{python}
plot

@@

```{python}
summary

## Explore the data

I plan to build a Shiny app to help compare trends across datasets (using Shinylive – built on Shiny for Python - ).

Here is a test app just so I can test the wiring of how such an app works.

from shiny import *
from plotnine import *
from pyodide.http import open_url
import pandas as pd
import numpy as np

#  ---------------------------------
#  ---------- SECTION 2: Read in data -------------
#  ----------------------------------

#  ---------------------------------
#  ---------- SECTION 3: Specifications -------------
#  ----------------------------------

# These parameters are not controlled in the app.
reference_col = 'Year'
group_by_col = 'Entity'

tolerance = 1
outlier_cut_off_upper = None

@@

app_ui = ui.page_fluid(
    {"class": "p-4"},
    ui.row(
        ui.column(
            4,
            panel_box(
                ui.input_slider("reference_vals", "Reference years", 1975, 2020, value=[1990,2018]),
                ui.input_numeric(
                            "max_dist_from_refs",
                            "=/- (years)",
                            4,
                        ),
                ui.input_numeric(
                            "min_dist_between",
                            "Min distance between observations (years)",
                            23,
                        ),
                ui.input_selectize(
                  "metrics", "Metrics", all_vars, multiple=True
                  ),
                ui.input_checkbox("all_metrics_requirement", "Show only countries with data for all metrics", True),
                ui.input_action_button(
                    "download", "Download data and image", class_="btn-primary w-100"
                ),
            ),
        ),
        ui.column(
            8,
            ui.output_plot("example_plot")
            ),
    ),
)

def server(input, output, session):
    @output
    @render.plot
    def example_plot():

      #  ---------------------------------
      #  ---------- SECTION 4: Run data functions and produce plot -------------
      #  ----------------------------------

      def prep_and_merge_multiple_vars(data_to_match, reference_vals, max_dist_from_refs, min_dist_between, reference_col, group_by_col):

        ref_matches = pd.concat(matches, keys=source_var_keys)

        # Tidy up indexes
        ref_matches = ref_matches.reset_index()

        ref_matches = ref_matches.drop('level_1', axis=1)

        ref_matches = ref_matches\
          .rename(columns={"level_0": "source_var"})

        return ref_matches

      def prep_plot_and_tables(
        data_to_match,
        reference_vals, 
        max_dist_from_refs,
        min_dist_between,
        reference_col,
        group_by_col,
        tolerance,
        outlier_cut_off_upper,
        all_metrics_requirement):

        # Other spec:
        region_col = 'region_alt'

        # ---- Prep data ---------

        plot_data = prep_and_merge_multiple_vars(
          data_to_match = data_to_match,
          reference_vals = reference_vals, 
          max_dist_from_refs = max_dist_from_refs,
          min_dist_between = min_dist_between,
          reference_col = reference_col,
          group_by_col = group_by_col
        )

        # ----- Prep plot ------

        # Set the coordinates of the shaded ribbon
        shaded_coord = pd.DataFrame({'x': [x_min_floor, x_max_ceiling, x_max_ceiling, x_min_floor], 
          'y': [x_min_floor-tolerance, x_max_ceiling-tolerance, x_max_ceiling+tolerance , x_min_floor+tolerance]})

        plot = (ggplot(plot_data
        , aes(x_axis, y_axis, alpha=0.5))
        + geom_point(aes(color=region_col, size='avg_pop', alpha = 0.6))
        + facet_wrap('~ source_var')
        + geom_polygon(aes(x='x', y='y', alpha = 0.2), data=shaded_coord)
        + theme_light()
        + theme(legend_position='none')
        )

        # ---- Prep summary table ----

        # ----- Return outputs ------
        plot_and_tables = {
          "table_data": plot_data,
          "plot": plot,
          "summary": df_summary
        }

        return plot_and_tables

      plot_and_tables = prep_plot_and_tables(
        data_to_match = data_to_match,
        reference_vals = list(input.reference_vals()), 
        max_dist_from_refs = input.max_dist_from_refs(),
        min_dist_between = input.min_dist_between(),
        reference_col = reference_col,
        group_by_col = group_by_col,
        tolerance = tolerance,
        outlier_cut_off_upper = outlier_cut_off_upper,
        all_metrics_requirement = input.all_metrics_requirement()
        )

      plot = plot_and_tables['plot']

@@

>>>>

---
title: "Appendix 1.F: Analysing global trends in within-country inequality"
format: html
warning: false
filters:
  - shinylive
sidebar: false

---

:::{.column-page}

## file: app.py

import jinja2
from shiny import *
import pandas as pd
from pyodide.http import open_url
import numpy as np
from plotnine import *

def prep_data(
      data_to_match,
      reference_vals,
      max_dist_from_refs,
      min_dist_between,
      all_metrics_requirement,
      reference_col,
      group_by_col,
      tolerance,
      outlier_cut_off_upper):

  plot_data = pd.concat(matches, keys=source_var_keys)

  # Tidy up indexes
  plot_data = plot_data.reset_index()

  plot_data = plot_data.drop('level_1', axis=1)

  plot_data = plot_data\
    .rename(columns={"level_0": "source_var"})

  #  PREP SUMMARY TABLE

  return plot_data, df_summary

#  ---------------------------------
#  ---------- SECTION 2: Specifications -------------
#  ----------------------------------
# These parameters are not controlled in the app.

reference_col = 'Year'
group_by_col = 'Entity'

tolerance = 1
outlier_cut_off_upper = None
region_col = 'region_alt'

#  ---------------------------------
#  ---------- SECTION 3: Read in data -------------
#  ----------------------------------
# Note when testing outside of shinylive, this needs to be adjusted to a normal pd_read, without the pyodide step.

#  ---------------------------------
#  ---------- SECTION 4: Gather data into dictionary and collete varnames -------------
#  ----------------------------------

#  ---------------------------------
#  ---------------------------------

app_ui = ui.page_fluid(
    {"class": "p-4"},
    ui.row(
        ui.column(
            4,
            panel_box(
                ui.input_slider("reference_vals", "Reference years", 1975, 2020, value=[1990,2018]),
                ui.input_numeric(
                            "max_dist_from_refs",
                            "=/- (years)",
                            4,
                        ),
                ui.input_numeric(
                            "min_dist_between",
                            "Min distance between observations (years)",
                            23,
                        ),
                ui.input_selectize(
                  "metrics", "Metrics", all_vars, multiple=True
                  ),
                ui.input_checkbox("all_metrics_requirement", "Show only countries with data for all metrics", True),
                ui.input_action_button(
                    "download", "Download data and image", class_="btn-primary w-100"
                ),
            ),
        ),
        ui.column(
            8,
            ui.output_plot("plot"),
            ui.output_table("table")
            ),
    ),
)

def server(input, output, session):

    @reactive.Calc
    def reactive_data():

      reference_vals = list(input.reference_vals())

      max_dist_from_refs = input.max_dist_from_refs()
      min_dist_between = input.min_dist_between()
      all_metrics_requirement = input.all_metrics_requirement()

      prepped_data = prep_data(
        data_to_match = data_to_match,
        reference_vals = reference_vals,
        max_dist_from_refs = max_dist_from_refs,
        min_dist_between = min_dist_between,
        all_metrics_requirement = all_metrics_requirement,
        reference_col = reference_col,
        group_by_col = group_by_col,
        tolerance = tolerance,
        outlier_cut_off_upper = outlier_cut_off_upper)

      return prepped_data

    @output
    @render.plot
    def plot():

        reference_vals = list(input.reference_vals())

        prepped_data = reactive_data()

        plot_data = prepped_data[0]

        plot = (ggplot(plot_data
        , aes(x_axis, y_axis, alpha=0.5))
        + geom_point(aes(color=region_col, size='avg_pop', alpha = 0.6))
        + geom_abline(intercept=0,slope=1) 
        + facet_wrap('~ source_var')
        + theme_light()
        + theme(legend_position='none')
        )

    @output
    @render.table
    def table():

        prepped_data = reactive_data()

        summary = prepped_data[1].reset_index()

        return summary

## file: requirements.txt
Jinja2

@@

:::

TESTING!!!
These are the same sections as above, to be used for testing.

from plotnine import *
import plotly.express as px
import pandas as pd
import numpy as np

@@

# Specify the tolerances for each kind of metric
def return_tolerance_for_var(var):

  if 'Gini' in var:
    tolerance = 1

  elif 'Top_10_share' in var:
    tolerance = 1

  elif 'Bottom_50_share' in var:
    tolerance = 1

  else:
    tolerance = None

  return tolerance

def prep_data(
      data_to_match,
      reference_vals,
      max_dist_from_refs,
      min_dist_between,
      all_metrics_requirement,
      reference_col,
      group_by_col,
      outlier_cut_off_upper):

@@

@@

    tolerance = return_tolerance_for_var(
      data_to_match["var_names"][i]
    )

@@

  plot_data = pd.concat(matches, keys=source_var_keys)

  # Tidy up indexes
  plot_data = plot_data.reset_index()

  plot_data = plot_data.drop('level_1', axis=1)

  plot_data = plot_data\
    .rename(columns={"level_0": "source_var"})

@@

@@

  #  PREP SUMMARY TABLE

@@

  return plot_data, df_summary

@@

#  ---------------------------------
#  ---------- SECTION 2: Specifications -------------
#  ----------------------------------
# These parameters are not controlled in the app.

reference_col = 'Year'
group_by_col = 'Entity'

outlier_cut_off_upper = None
region_col = 'region_alt'

@@

#  ---------------------------------
#  ---------- SECTION 3: Read in data -------------
#  ----------------------------------

@@

#  ---------------------------------
#  ---------- SECTION 4: Gather data into dictionary and collete varnames -------------
#  ----------------------------------

@@

@@

We don't run this reactive step in the testing. We just run the prep data function in the plot and table functions below.

```{python}
# def reactive_data():

#   # For testing we define manually, rather than picked up from the app inputs
#   data_to_match = prep_selected_vars(input.metrics())

#   reference_vals = list(input.reference_vals())

#   max_dist_from_refs = input.max_dist_from_refs()
#   min_dist_between = input.min_dist_between()
#   all_metrics_requirement = input.all_metrics_requirement()

#   prepped_data = prep_data(
#       data_to_match = data_to_match,
#       reference_vals = reference_vals,
#       max_dist_from_refs = max_dist_from_refs,
#       min_dist_between = min_dist_between,
#       all_metrics_requirement = all_metrics_requirement,
#       reference_col = reference_col,
#       group_by_col = group_by_col,
#       tolerance = tolerance,
#       outlier_cut_off_upper = outlier_cut_off_upper)

#   return prepped_data

def interactive_plot(
  selected_vars,
  reference_vals,
  max_dist_from_refs,
  min_dist_between,
  all_metrics_requirement):

  data_to_match = prep_selected_vars(selected_vars)

  prepped_data = prep_data(
      data_to_match = data_to_match,
      reference_vals = reference_vals,
      max_dist_from_refs = max_dist_from_refs,
      min_dist_between = min_dist_between,
      all_metrics_requirement = all_metrics_requirement,
      reference_col = reference_col,
      group_by_col = group_by_col,
      outlier_cut_off_upper = outlier_cut_off_upper)

  plot_data = prepped_data[0]

  fig = px.scatter(
    plot_data, 
    x=f'value{reference_vals[0]}', 
    y=f'value{reference_vals[1]}',
    color=region_col, 
    facet_col='source_var', 
    facet_col_wrap=1, 
    size='avg_pop',
    hover_name=group_by_col,
    hover_data={
      f'value{reference_vals[0]}':':.2f',
      f'{reference_col}{reference_vals[0]}':True,
      f'value{reference_vals[1]}':':.2f',
      f'{reference_col}{reference_vals[1]}':True,
      region_col: True,
      'avg_pop': False,
      'source_var': False},
      height=1000
    )

  fig.update_layout(legend=dict(
      orientation="h",
      yanchor="bottom",
      y=1.02,
      xanchor="right",
      x=1
  ))

  fig.for_each_annotation(lambda a: a.update(text=a.text.split("=",1)[1]))

  fig.update_layout(
      legend_title=""
  )

  return fig

```{python}
def table_test(selected_vars,
  reference_vals,
  max_dist_from_refs,
  min_dist_between,
  all_metrics_requirement):

  data_to_match = prep_selected_vars(selected_vars)

  prepped_data = prep_data(
    data_to_match = data_to_match,
    reference_vals = reference_vals,
    max_dist_from_refs = max_dist_from_refs,
    min_dist_between = min_dist_between,
    all_metrics_requirement = all_metrics_requirement,
    reference_col = reference_col,
    group_by_col = group_by_col,
    outlier_cut_off_upper = outlier_cut_off_upper)

  summary = prepped_data[1].reset_index()

  return summary

fig = interactive_plot(
  selected_vars = [
      'WID: Top_10_share_pretax',
      'PIP: Top_10_share'],
  reference_vals = [1990,2015],
  max_dist_from_refs = 5,
  min_dist_between = 10,
  all_metrics_requirement = True
)

fig.show()

summary_table = table_test(
  selected_vars = [
      'WID: Top_10_share_pretax',
      'PIP: Top_10_share'],
  reference_vals = [1990,2015],
  max_dist_from_refs = 5,
  min_dist_between = 10,
  all_metrics_requirement = True
)

summary_table

### 2023-02-09 PhD_pages/paper_pages/paper1.qmd
Paper 1: reorganise app and initial drafting of paper

What has happened to within-country inequality globally?

>>>>

What has happened to inequality within countries over the past generation?

# Introduction

- A range of motives for taking a global view: SDGs; globalization – trade; ....

- We don't have a clear global view.    
  [UN summary of inequality trends](https://www.un.org/en/un75/inequality-bridging-divide)

- Why is it that we don't have a clear view?

    1) Most academic research and commentary has focussed on high income countries (refs), especially the US. There are many reasons why we should not expect trends in inequality in these countries to reflect broader global trends.

    2) Hampered by the available data
        - **Coverage** – for instance no surveys.
        - **Comparability of data** – Some secondary databases with global coverage mixes very heterogenous datapoints (WIID) (Find Atkinson and Brandolini ref from Handbook...). Those with a greater degree of harmonization tend to lack global coverage (LIS, EU-SILC, SEDLAC, until recently PIP)
        - **Some trade off between the two is inevitable** – There is no global survey of incomes. Countries run surveys based on their own needs and capacities. As such they differ. Saying something global means accepting a degree of heterogenity in the data. Similarly, definitions of income for tax purposes vary, impacting the comparability of tax data.

    3) Contradictory global views are presented by different global databases.
        - The data situation has improved drastically: thanks to harmonized databases with global coverage. In particular:
            - PIP – recently expanded to include high income countries
            - WID – Transition from WTID, and the DINA methodology, global coverage.
            - LIS: some data from middle and low income countries.
        - But the 'take home' messages from the institutions producing these databases offer very contradictory views.
            - "In recent decades, income inequality 
has increased in nearly all countries, 
but at different speeds, suggesting 
that institutions and policies matter in shaping inequality." (WIR 2018)
            - "[Across a range of approaches] aver-
age  inequality  within  countries  increased   in  the  1990s,  but  has  been  falling  in  recent years according to most specifications." (World Bank, 2016)

        - You can see this in the latest data from these two sources, using the metric typically favoured by each: Top 1% from WID vs Gini from PIP.

        - These different messages are based on a mix of many different methodological differences:
            - different metrics, sensitive to different parts of the distribution; 
            - Different welfare concepts; 
            - Different underlying data sources and adjusmtents to the top (or lack thereof);
            - Different coverage or time periods being analysed.

# Literature review – Global studies of inequality

- Atkinson and Brandolini (2010) argue that global inequality analyses in particular need to consider both absol ute  and relative differences.

# Data review Why is our understanding of inequality around the world currently limited

- **A global view is hampered by the available data**.

- **But databases have improved substantially recently**
Thanks to improvements and expansion of key global harmonized databases, we are now in a much stronger position to take a global view on inequality within countries.

There are range of databases and a full review is out of the scope of this paper. Focus on the three databases of harmonized distributional data providing global or near-global coverage.

- **PIP** 
    - Survey data
    - Mix of disposable income and consumption data (definitions vary)
    - Per capita
    – Coverage: now includes high income countries (Add a box/visualization with coverage). Global coverage from 1990 onwards.
    - A range of inequality indices provided 'as standard', but the whole distribution may be queried via API. Either microdata or, in a minority of cases, a parametric distribution fitted to 'grouped data' (e.g. income deciles)

- **WID** 
    – Survey data, tax data, National accounts
    - Coverage much greater for pre-tax national income. But post-tax concepts also available for a subset of observations.
    - Greatest coverage for individuals – equal split where underlying data was available for couples.
    - Coverage 
    - Estimates of the whole distribution Generalized Pareto Interpolation. Though longer-run coverage is higher for pretax top 1% share (reliance on tax data)

- **LIS**
    - Offers less coverage, but is notable for providing access to microdata.
